A Large-Scale Study of Polymorphism in R

### 2019-04-06 paper/main.tex
update

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{20\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
\newcommand{\PERCSCALARMONO}{37.6\%\xspace}

\newcommand{\LZEROPERCPOLY}{17\%\xspace}
\newcommand{\LONEPERCPOLY}{20.4\%\xspace}
\newcommand{\LTWOPERCPOLY}{23\%\xspace}

@@

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
\begin{table}[ht]
\label{tab:L0top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 69806 & 26.3\% \\ 
  \l\ & 55311 & 20.8\% \\ 
  \C & 27789 & 10.5\% \\ 
  \L & 26230 & 9.9\% \\ 
  \sF & 16806 & 6.3\% \\ 
  \sN & 13976 & 5.3\% \\ 
  \ANY & 9125 & 3.4\% \\ 
  \I & 8971 & 3.4\% \\ 
  \sS & 4748 & 1.8\% \\ 
  \D, \I & 3256 & 1.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\end{table}

We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
Recall that in {\bf L0}, data frames will appear as lists.
The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, and if this metaprogramming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming; we will discuss it in a little more detail shortly.

One of our chief goals is to generate useful function signatures.
After tracing and capturing unique signatures from function call sites, recall from Section~\ref{sec:autosub} that we collapse some of the more obvious ones together.
After this, we are curious to know: {\it how many functions have more than one signature?}
Figure~\ref{fig:L0funcounts} has the answer with a breakdown by function.

@@

Here, we see that the vast majority of functions are monomorphic in L0.
That said, there are a few functions with a disproportionately large number of signatures.
One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl}).
Many of the metaprogramming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
This usability maps to polymorphism in our analysis, so it's no surprise that a Tidyverse function features so prominently here.

Another function with many signatures is the data.table package's [ function, which is overloaded for objects with data.table class.
This function has 16 formal arguments, and call configurations vary widely.
Essentially, [ is {\it the} main function for data.tables, and performs very different tasks based on which arguments are provided.
Two wildly different uses are:
\code{DT[i, j, by]}, where columns \code{i} and \code{j} are accessed and grouped by \code{by}, and \code{DT[x == "some" & y > 42]}, which selects all rows who's \code{x} column is the string \code{"some"} and who's \code{y} column is greater than \code{42}.
Given that, we find it unsurprising that the function is so polymorphic.

To get an idea of the most common forms of argument-level polymorphism, we refer toTable~\ref{tab:L0toppoly}.

\begin{table}[ht]\label{tab:L0toppoly}\centering
\begin{tabular}{lrr}  \hline
Signature & Count & \% of All Signatures \\
\hline
  \D, \l & 2799 & 1\% \\ 
  \D, \sN & 2290 & 0.9\% \\ 
  \C, \sN & 1527 & 0.6\% \\ 
  \l, \sN & 1193 & 0.4\% \\ 
  \C, \D & 858 & 0.3\% \\ 
  \D, \sF & 413 & 0.2\% \\ 
  \C, \I & 382 & 0.1\% \\ 
  \I, \sN & 378 & 0.1\% \\ 
  \C, \l & 371 & 0.1\% \\ 
  \C, \sF & 334 & 0.1\% \\ 
\end{tabular}
\caption{Top polymorphic argument signatures for the L0 type system.}
\end{table}

@@

	\item \D, \sF: this signature appears more strange than it is.
	Many models, processes, simulations, etc. that are prevalent throughout R are are {\it parameterized}.
	Some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
	Others, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are either parameters for complex data generation processes (\D), or a function to be used during the generation (\sF).
	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers.

@@

\begin{table}[ht]\label{tab:L0argcounts}\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 250K & 94\% \\ 
  Total Seen & 266K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

From Table~\ref{tab:L0argcounts} we see that, with our definition of subtyping on L0 and other simplifications we made, an overwhelming amount of argument polymorphism is accounted for.
We also saw in Table~\ref{tab:L0toppoly} that each of the most frequently occurring argument signatures had some explanation.
That said, L0 is not without limitations.

The main issues with L0 stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
The next type system we will explore, L1, tackles these issues by collecting more information to build more precise types.
While the L1 types are less ``in-tune'' with current R implementations, one can imagine designing an all-new R runtime which may benefit from the more specific information in L1, e.g. by not needing to vectorize scalar values, or designing a static analysis which uses precise list types to give programmers better information in an IDE.
L1 is discussed next.

@@

Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\label{tab:L1top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 48815 & 18.4\% \\ 
  \D & 27367 & 10.3\% \\ 
  \sL & 24284 & 9.2\% \\ 
  \sC & 22025 & 8.3\% \\ 
  \l & 20062 & 7.6\% \\ 
  \sF & 18480 & 7\% \\ 
  \M{D} & 14315 & 5.4\% \\ 
  \sN & 14016 & 5.3\% \\ 
  \ANY & 9093 & 3.4\% \\ 
  \C & 7731 & 2.9\% \\
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\end{table}

From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
We see that scalars actually figure quite prominently, which is encouraging, supporting our suggestion of an explicit scalar annotation.
We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.

@@

We see that we increased the polymorphism with our transition to L1, despite the monomorphic nature of the most frequently appearing argument signatures.
As we were in L0, we are curious to know how argument-level polymorphism looks in L1.
Refer to Table~\ref{tab:toppolyL1}.

\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \M{D} & 1440 & 0.5\% \\ 
  \l, \lT{D} & 525 & 0.2\% \\ 
  \M{D}, \M{I} & 383 & 0.1\% \\ 
  \sC, \sD & 382 & 0.1\% \\ 
  \I, \sD & 370 & 0.1\% \\ 
  \l, \lT{list} & 337 & 0.1\% \\ 
  \df, \M{D} & 305 & 0.1\% \\ 
  \C, \D & 294 & 0.1\% \\ 
  \sD, \sF & 290 & 0.1\% \\ 
  \sC, \sF & 273 & 0.1\% \\
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}
\end{table}

@@

	\item \sD, \sF and \sC, \sF: these were thoroughly discussed in the context of L0.

@@

\begin{table}[ht]
\label{tab:argcountsL1}
\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 252K & 94.9\% \\ 
  Total Seen & 265K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L1.}
\end{table}

In spite of drastically increasing the precision of our types, we are still happily only dealing with a small amount of polymorphism.
This suggests that with a good notion of subtyping, precise type annotations won't cause an undue increase in the polymorphism of the language.

@@

Our suggestion to add an explicit scalar type is not unfounded, though it is not compatible with current R implementations.
We saw in this section that strictly scalar values appear quite frequently, making up some of the most frequently occurring argument signatures.
In fact, \PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
This suggests that such a distinction would more accurately capture current use of the language, and would also be beneficial to future implementations which can perform optimizations based on the type (e.g., not vectorizing scalar-only values).

@@

Throughout this section, we have seen that even though we substantially increased the precision of our type annotations, we have not increased the amount of polymorphism that we observe.
The transition from L0 to L1 revealed some interesting patterns that L1 cannot quite capture---for these, we will need to appeal to the class and attributes of values.
The next type system we consider is L2, and is designed to achieve this.

@@

\begin{table}[ht]
\label{tab:L2top10arg}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \sD & 46938 & 17.6\% \\ 
  \sL & 24113 & 9\% \\ 
  \sC & 21387 & 8\% \\ 
  \D & 20900 & 7.8\% \\ 
  \sF & 14556 & 5.5\% \\ 
  \sN & 14016 & 5.3\% \\ 
  \ANY & 9093 & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 6974 & 2.5\% \\ 
  \C& 6939 & 2.6\% \\ 
  \attrclass{\l}{}{names} & 6137 & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\end{table}

As you can see, the most frequent signatures in {\bf L1} are represented here.
The inclusion of class and attribute information has changed the counts slightly.
One tidbit of interest here is that monomorphic named lists are more common than their unnamed counterparts.
The lack of classes is explained by our elimination of builtin classes, and it's doubtful that a user-defined class would be so pervasive as to feature in this breakdown.

Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
Now, {\bf L2} builds directly off of {\bf L1}, and as such at the argument level the type-based polymorphism shouldn't change.
For this account of polymorphism, we deem an argument to be polymorphic if it was polymorphic in type, class, or attribute pattern.
Consider Figure~\ref{fig:L2funcounts}.

@@

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Apr  5 09:50:59 2019
\begin{table}[ht]
\label{tab:toppolyL2}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \attrclass{\D}{}{dim} & 553 & 0.2\% \\ 
  \sC, \sD & 336 & 0.1\% \\ 
  \I, \sD & 320 & 0.1\% \\ 
  \attrclass{\l}{}{names}, \attrclass{\D}{}{names} & 269 & 0.1\% \\ 
  \D, \attrclass{\D}{}{dim, dimnames} & 218 & 0\% \\ 
  \sD, \sF & 173 & 0\% \\ 
  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 167 & 0\% \\ 
  \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 155 & 0\% \\ 
  \C, \I & 149 & 0\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in {\bf L2}.}
\end{table}

When comparing this table with Table~\ref{tab:toppolyL1}, we see that the distribution of signature frequencies is more uniform, as attributes and classes differentiated some signatures which were not distinct in {\bf L1}.

@@

\begin{table}[ht]
\label{tab:argcountsL2}
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 248K & 93.4\% \\ 
  Total Seen & 265K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in {\bf L2}.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\label{tab:packagepolysize}
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  knitLatex & 4 & 41 & \LaTeX \xspace generator from R data \\ 
  magrittr & 17 & 126 & Tidyverse pipe suite \\ 
  proto & 5 & 30 & Prototype object-based programming \\ 
  gnlm & 10 & 53 & Regression fit package \\ 
  Formula & 17 & 82 & ?? \\ 
   \hline
\end{tabular}
\caption{5 most polymorphic packages (in {\bf L0}).}
\end{table}

Now, what makes these so polymorphic?

\AT{TODO if time: fix to score per fun, then tally per package.}

@@

	\item magrittr: we discussed magrittr earlier in the context of its pipe \%$>$\% function. 
	It's not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g. \%$<>$\% which pipes as normal and then assigns to the left-hand argument.

	\item proto: most of the polymorphism in this package comes from its implementation of assignment ({\tt \$<-}) for proto class objects.
	It's a small, 3 argument function which is exceedingly polymorphic, with over 20 unique call signatures.
	As theoretically any R object can have a proto class, it stands to reason that this function ought to be quite polymorphic indeed.

@@

To determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
There may also be other interesting per-function trends which our current analysis does not pick up on.
We aim to make our data available for this purpose.

@@

>>>>

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{20\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
\newcommand{\PERCSCALARMONO}{38.2\%\xspace}

\newcommand{\LZEROPERCPOLY}{19\%\xspace}
\newcommand{\LONEPERCPOLY}{18.1\%\xspace}
\newcommand{\LTWOPERCPOLY}{21.3\%\xspace}

@@

Across all of the arguments that we've observed, the vast majority are monomorphic.
Table~\ref{tab:L0top10arg} shows the breakdown of the 10 most frequently occurring argument signatures.
\begin{table}[ht]
\label{tab:L0top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \D & 269K & 29.7\% \\ 
  \l & 193K & 21.3\% \\ 
  \C & 96K & 10.6\% \\ 
  \L & 83K & 9.2\% \\ 
  \sF & 60K & 6.6\% \\ 
  \sN & 47K & 5.2\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \I & 30K & 3.2\% \\ 
  \sS & 16K & 1.7\% \\ 
  \D, \l & 10K & 1.1\% \\ 
   \hline
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L0}.}
\end{table}

We see that doubles are the most common argument type, which is not entirely surprising as doubles are the backbone of any numeric data.
Next is lists, which seems reasonable given that lists are often used to store the aforementioned data.
Recall that in {\bf L0}, data frames will appear as lists.
The \ANY type is not an R type per se: we ascribe such a type to arguments which were {\it always} unevaluated across all calls.
As it happens, many R functions metaprogram on some of their arguments, and this metaprogramming does not force the argument promise, and if this metaprogramming is always performed in every recorded call we will ascribe the \ANY type, as we cannot say what the argument {\it is} without possibly forcing a promise.
The magrittr package's pipe \%$>$\% function is an example of a function which performs a lot of metaprogramming; we will discuss it in a little more detail shortly.
Interestingly, a polymorphic attribute signature \D, \l is among the top 10 overall---this is the vector and list polymorphism (shown in Figure~\ref{fig:listvec}) made manifest.

Now that we've seen the most common attribute-level signatures, we will turn our attention to function-level polymorphism.
Indeed, one of our goals is to generate useful function signatures, so we ask: {\it how many functions have more than one recorded signature?}
Figure~\ref{fig:L0funcounts} has the answer, with a breakdown by function.

@@

Here, we see that the vast majority of functions are monomorphic in L0.
That said, there are a few functions with a disproportionately large number of signatures.
One such function is the magrittr package's pipe \%$>$\%, which is used to implement a kind of currying (as per discussion in Section~\ref{sec:Impl})---here, \%$>$\% has nearly 100 unique signatures!
Many of the metaprogramming functions in the Tidyverse are quite sophisticated, and are designed with the explicit goal of being very usable.
This usability maps to polymorphism in our analysis, so it's no surprise that a Tidyverse function features so prominently here.

Another function with many signatures is the data.table package's [ function, which is a reimplementation of the square bracket operator for data.tables.
This function has 16 formal arguments, and call configurations vary widely.
Essentially, [ is {\it the} main function for data.tables, and performs very different tasks based on which arguments are provided.
Two wildly different uses are:
\code{DT[i, j, by]}, where columns \code{i} and \code{j} are accessed and grouped by \code{by}, and \code{DT[x == "some" & y > 42]}, which selects all rows who's \code{x} column is the string \code{"some"} and who's \code{y} column is greater than \code{42}.
Given that, we find it unsurprising that the function is so polymorphic.

Having had a taste for function-level polymorphism, we will now explore common patterns in argument-level polymorphism.
For this, refer to Table~\ref{tab:L0toppoly}.

\begin{table}[ht]\label{tab:L0toppoly}\centering
\begin{tabular}{lrr}  \hline
Signature & Count & \% of All Signatures \\
\hline
  \D, \l & 9566 & 1.1\% \\ 
  \D, \sN & 8801 & 1\% \\ 
  \C, \sN & 5590 & 0.6\% \\ 
  \l, \sN & 5016 & 0.6\% \\ 
  \C, \D & 2887 & 0.3\% \\ 
  \D, \sF & 1608 & 0.2\% \\ 
  \l, \sF & 1345 & 0.1\% \\ 
  \C, \l & 1283 & 0.1\% \\ 
  \C, \sF & 1168 & 0.1\% \\ 
  \C, \I & 1104 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures for the L0 type system.}
\end{table}

@@

	\item \D, \sF: this signature appears more strange than it is.
	Many models, processes, simulations, etc. that are prevalent throughout R have parameters to fine-tune their behavior.
	This pattern manifests itself here: some arguments with the \D, \sF signature are either parameters for a function (\D), or the function itself already constructed (\sF).
	Others, such as the {\tt br} argument of the {\tt rtree} function in the {\tt ape} package, are either parameters for complex data generation processes (\D), or a function to be used during the generation (\sF).
	It would appear that this pattern is used to make complex simulation, modelling, or data generation functions more flexible for programmers by allowing them to supply parameters or the parameterized function itself.

@@

\begin{table}[ht]\label{tab:L0argcounts}\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 852K & 94\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

From Table~\ref{tab:L0argcounts} we see that, with our definition of subtyping on L0 and other simplifications we made, an overwhelming amount of argument polymorphism is accounted for.
We also saw in Table~\ref{tab:L0toppoly} that each of the most frequently occurring argument signatures had some explanation.

That said, L0 is not without limitations, and the main issues stem from a lack of adequate information, as the type system doesn't capture precise list types, and does not distinguish scalars from vectors.
The next type system we will explore, L1, tackles these issues by collecting more information to build more precise types.
While the L1 types are less ``in-tune'' with current R implementations, one can imagine designing an all-new R runtime which may benefit from the more specific information in L1, e.g. by not needing to vectorize scalar values, or designing a static analysis which uses precise list types to give programmers better information in an IDE.
L1 is discussed next.

@@

Let us begin as we did in L0 by looking at the breakdown of argument signatures in L1.
Consult Table~\ref{tab:L1top10arg}.
\begin{table}[ht]
\label{tab:L1top10arg}
\centering
\begin{tabular}{lrr}
  \hline
 Argument Signature & Count & Percentage \\ 
  \hline
  \sD & 179K & 19.7\% \\ 
  \D & 97K & 10.7\% \\ 
  \sL & 77K & 8.6\% \\ 
  \sC & 75K & 8.3\% \\ 
  \l & 67K & 7.4\% \\ 
  \sF & 61K & 6.7\% \\ 
  \sN & 50K & 5.5\% \\ 
  \M{D} & 49K & 5.5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \df & 29K & 3.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in L1.}
\end{table}

From Table~\ref{tab:L1top10arg}, we see that all of the most frequent argument signatures are monomorphic even though we increased the precision of the types.
We see that scalars actually figure quite prominently, which is encouraging: this supports an idea that an explicit scalar annotation or type may be useful {\it for programmers}---indeed, such an annotation would be helpful for compiler optimizations (e.g., scalar-only values need not have vectors allocated to store them).
We also see that matrices (\M{D}) are quite popular, a testament to the versatility of the data structure, as they can easily be converted to vectors and data frames.

@@

We see that we barely increased the polymorphism with our transition to L1, despite the far richer type information.
The biggest outlier here is indeed \%$>$\%, as was the case in {\bf L0}.
Otherwise, the shape is very similar.
In short: even though {\L1} types are more expressive, the breakdown of per-function polymorphism is largely unchanged from {\bf L0}.

As we were in L0, we are curious to know how argument-level polymorphism looks in L1.
Refer to Table~\ref{tab:toppolyL1}.

\begin{table}[ht]
\label{tab:toppolyL1}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \M{D} & 4848 & 0.5\% \\ 
  \l, \lT{D} & 1843 & 0.2\% \\ 
  \sC, \sD & 1590 & 0.2\% \\ 
  \M{D}, \M{I} & 1502 & 0.2\% \\ 
  \l, \lT{list} & 1199 & 0.1\% \\ 
  \df, \M{D} & 1180 & 0.1\% \\ 
  \C, \D & 1019 & 0.1\% \\ 
  \sD, \sF & 1005 & 0.1\% \\ 
  \I, \sD & 997 & 0.1\% \\ 
  \sC, \sF & 991 & 0.1\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in L1.}
\end{table}

@@

	\item \sD, \sF and \sC, \sF: these were discussed in the context of L0.

@@

\begin{table}[ht]
\label{tab:argcountsL1}
\centering
\begin{tabular}{lrl}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
Monomorphic & 864K & 95.3\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L1.}
\end{table}

In spite of drastically increasing the precision of our types, we have actually {\it reduced} the amount of polymorphism.
This suggests that with a good notion of subtyping, precise type annotations won't cause an undue increase in the polymorphism of the language.

@@

It seems that there is a place for a scalar annotation in R.
We saw in this section that strictly scalar values appear quite frequently, making up some of the most frequently occurring argument signatures.
In fact, \PERCSCALARMONO of arguments have a strictly scalar monomorphic signature.
This suggests that such a distinction would more accurately capture current use of the language, and would also be beneficial to future implementations which can perform optimizations based on the type (e.g., not vectorizing scalar-only values).

@@

Throughout this section, we have seen that even though we substantially increased the precision of our type annotations, we have not increased the amount of polymorphism that we observe.
The transition from L0 to L1 revealed some interesting patterns that L1 cannot quite capture---for these, we will need to appeal to the class and attributes of values, which are incorporated into {\bf L2}, discussed next.

@@

\begin{table}[ht]
\label{tab:L2top10arg}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
 \sD & 172K & 19\% \\ 
  \sL & 77K & 8.5\% \\ 
  \D & 77K & 8.4\% \\ 
  \sC & 73K & 8\% \\ 
  \sN & 50K & 5.5\% \\ 
  \sF & 45K & 5\% \\ 
  \ANY & 31K & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 25K & 2.6\% \\ 
  \C & 25K & 2.6\% \\ 
   \attrclass{\l}{}{names, row.names}  & 21K & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\end{table}

As you can see, the most frequent signatures in {\bf L1} are represented here.
The inclusion of class and attribute information has changed the counts slightly.
Not pictures here, but interesting: named lists are more frequent than their unnamed counterparts.
No classes are represented here since (1) we eliminated builtin classes, as we've used them to construct types already, and (2) most other classes are specific to some package or set of packages---it's doubtful that such a class would be so pervasive as to feature in this breakdown.

Now that we have a sense for argument-level type signatures in {\bf L2}, we will see just how many functions are polymorphic.
Consider Figure~\ref{fig:L2funcounts}.

@@

Not pictured here are two outlier values: the summary and print functions of the base R package.
Summary is a generic function which is used to summarize various types of data, and print is used to convert values to string format and print them to the console.
These functions really {\it should} be outliers, as they are really operating over any type.
They feature so prominently here since many of the values which flowed into them had some amount of attribute or class information.

Overall, the amount of polymorphism only slightly increased, even though we added two new dimensions of type information.
Again, far richer types do not increase the amount of polymorphism by significant amounts, perhaps a testament to the monomorphic nature of many R functions.

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Apr  5 09:50:59 2019
\begin{table}[ht]
\label{tab:toppolyL2}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \D, \attrclass{\D}{}{dim} & 1942 & 0.2\% \\ 
  \sC, \sD & 1431 & 0.2\% \\ 
  \attrclass{\l}{}{names}, \attrclass{\D}{}{names}  & 1046 & 0.1\% \\ 
  \I, \sD & 855 & 0.1\% \\ 
  \D, \attrclass{\D}{}{dim, dimnames} & 810 & 0\% \\ 
  \attrclass{\l}{}{names, row.names}, \attrclass{\D}{}{dim, dimnames} & 708 & 0.1\% \\ 
  \C, \D & 672 & 0\% \\ 
   \attrclass{\D}{}{names}, \attrclass{\D}{}{dim, dimnames} & 591 & 0\% \\ 
  \attrclass{\D}{}{dim}, \attrclass{\I}{}{dim}  & 445 & 0\% \\ 
  \sD, \attrclass{\sF}{}{srcref} & 445 & 0\% \\ 
   \hline
\end{tabular}
\caption{Top polymorphic argument signatures in {\bf L2}.}
\end{table}

When comparing this table with Table~\ref{tab:toppolyL1}, we see that by-and-large the same types are represented.
We note the {\tt names} attributes on the lists in row 3: these named lists are, in general, more common than their unnamed counterparts.
In the 10th signature, we see a function with attribute {\tt srcref}: functions with this attribute are for dealing with open files in the context of the srcfile set of functions in the base package.
As for why such a function would share an argument position with scalar double, there are other representations of open files, and numbers feature among them.

@@

\begin{table}[ht]
\label{tab:argcountsL2}
\centering
\begin{tabular}{lll}
  \hline
Signature Type & Count & Percentage \\ 
  \hline
  Monomorphic & 851K & 93.9\% \\ 
  Total Seen & 906K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in {\bf L2}.}
\end{table}

@@

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\label{tab:packagepolysize}
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  magrittr & 20 & 225 & Tidyverse pipe suite \\ 
  oshka & 4 & 38 & quoted expression expansion \\ 
  scatterD3 & 1 & 9 & create D3 (JS) scatter plots from R \\ 
  yaml & 4 & 33 & convert R to YAML \\ 
  knitLatex & 4 & 32 & \LaTeX \xspace generator from R data \\ 
   \hline
\end{tabular}
\caption{5 of the most polymorphic packages (in {\bf L0}).}
\end{table}

Now, what makes these so polymorphic?
The first package, magrittr, is one that we discussed earlier in the context of its pipe \%$>$\% function. 
It's not surprising that the package as a whole is quite polymorphic, as it implements {\it several} different pipe functions, e.g. \%$<>$\% which pipes as normal and then assigns to the left-hand argument.
In fact, all but knitLatex implement some form of metaprogramming:
oshka meta
\begin{enumerate}

@@

@@

Another note regarding function signatures is the {\tt ...} function argument.
Functions with such an argument can be passed an arbitrary number of extra arguments, see Figure~\ref{fig:dotdotdot} for a small example.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
f <- function(x, ...) {
  g(x, ...)
}

f(42, z=2, y=4) # in f, will perform g(42, z=2, y=4)

\end{lstlisting}}\caption{Simple ... example.}\label{fig:dotdotdot}\end{figure}

Our analysis ignores these extra arguments, instead flagging functions which have them.
If building an annotation for such a function, the ... argument is always ascribed the \l type.

Ultimately, to determine which kinds of whole function signatures would be useful, a more sophisticated analysis is required, which would need to develop a correlation between function arguments and return types.
There may also be other interesting per-function trends which our current analysis does not pick up on.
We aim to make our data available for this purpose.

@@

OUTLINE
